---
title: "Basic_Backbone"
author: "Robin Vandaele"
output: html_document
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of this tutorial is to show the basic functionality of the backbone function.
We first import the necessary libraries.

```{r, warning=FALSE, message=FALSE}
devtools::load_all() # load graphTDA
library("ggplot2") # plotting
```

We load and plot the toy data set that will be used for this example.

```{r, fig.width=3, fig.height=4}
df <- read.table("~/Documents/graphTDA/data/curvY.csv", header=FALSE, sep=",", col.names=c("x", "y"))
ggplot(df, aes(x=x, y=y)) +
  geom_point(size=3) +
  theme_bw() +
  coord_fixed()
```

The most important function for identifying backbones in graphs is, as you might guess, the 'backbone' function.
Let's illustrate its functionality by computing a backbone from a (Vietoris-)Rips graph constructed from the point cloud.

```{r}
BCB <- backbone(df, # point cloud data.frame/matrix, dist object, or a given graph
                type="rips", # type of proximity graph to construct (standard is knn)
                eps=10, # defining parameter for the rips graph
                assign=TRUE) # whether to assign nodes to branches in the backbone, clarified below
```

The constructed proximity graph is stored in the list entry 'G'.
As no vertex-valued function was provided to compute the pine, the boundary coefficients were computed and stored in the list entry 'f'.
Note the order of the vertices in the constructed graph may differ from the order of points in the data.
This needs to be taken into account for visualization (values in f are ordered according to nodes in the graph).
We can visualize the graph and boundary coefficients as follows.

```{r, fig.width=4, fig.height=4}
EG <- get_edges2D(df, BCB$G)
ggplot(df[names(V(BCB$G)),], aes(x=x, y=y)) +
  geom_segment(data=EG, aes(x=x1, y=y1, xend=x2, yend=y2), color='black', alpha=0.15) +
  geom_point(size=2, aes(col=BCB$f)) +
  scale_colour_gradientn(colours=topo.colors(7)) +
  labs(col="BC") +
  theme_bw() +
  coord_fixed()
```

The f=BC-pine is constructed as an intermediate representation for topological inference.
It is stored in the list entry 'pine', and can be visualized as follows.

```{r, fig.width=4, fig.height=4}
Epine <- get_edges2D(df, BCB$pine)
ggplot(df[names(V(BCB$G)),], aes(x=x, y=y)) +
  geom_segment(data=EG, aes(x = x1, y = y1, xend = x2, yend = y2), color="black", alpha=0.15) +
  geom_segment(data=Epine, aes(x=x1, y=y1, xend=x2, yend=y2), color="red", size=1) +
  geom_point(size=1, aes(col=BCB$f)) +
  scale_colour_gradientn(colours=topo.colors(7)) +
  labs(col="BC") +
  theme_bw() +
  coord_fixed()
```

The inferred backbone is stored in the list entry 'B'.
The option 'assign=TRUE' (standard is FALSE) results in an assignment of points to branches of B, i.e., maximal paths without multifurcations.
We can visualize the backbone B along with the branch assignment as follows.

```{r, fig.width=4, fig.height=4}
EBCB <- get_edges2D(df, BCB$B)
ggplot(df[names(V(BCB$G)),], aes(x=x, y=y)) +
  geom_segment(data=EG, aes(x=x1, y=y1, xend=x2, yend=y2), color="black", alpha=0.1) +
  geom_segment(data=Epine, aes(x=x1, y=y1, xend=x2, yend=y2), color="black", alpha=0.5, size=1) +
  geom_point(size=3, fill=BCB$col, alpha=0.5, pch=21) +
  geom_segment(data=EBCB, aes(x= x1, y=y1, xend=x2, yend=y2), col="black", size=2.75) +
  geom_segment(data=cbind(EBCB, as.factor(BCB$branch)),
               aes(x=x1, y=y1, xend=x2, yend=y2, col=as.factor(BCB$branch)), size=2) +
  geom_point(data=df[V(BCB$B)$name,], fill=BCB$col[V(BCB$B)$name], size=5, pch=21) +
  scale_color_manual(values=BCB$palette) +
  guides(col = guide_legend(title="Path")) +
  coord_fixed() +
  theme_bw()
```

Since no number of leaves was provided, an 'elbow' estimate was performed on the cost curve obtained by solve CLOF.
Although a tree with one leaf is technically not possible, we store the cost of the first 'optimal branch' to track the evolution of the cost.
This branch can be identified whenever three leaves are present in the backbone.
For 0 leaves, the cost function marks the maximal cost of a node in the graph if CLOF is solved for a vertex-valued function.
This is the case here, as the standard setting is to use the betweenness centrality of the pine for solving CLOF.
As this function is constant on leaves, the pine is prepruned, significantly decreasing the number of possible leaves for computational efficiency.
In this case, the last value of the cost curve will mark the increase that would be achieved when an arbitrary leaf would be added or extended.
The relative cost function is stored in the list entry 'cost', and can be visualized as follows.
Note that the absolute cost function can be obtained from the relative cost function and the global cost that is stored in the list entry 'full_cost'.

```{r, fig.width=6, fig.height=3}
ggplot(BCB$cost, aes(x=leaves, y=cost)) +
  geom_line(aes(group=component, col=component)) +
  geom_point(aes(group=component, col=component)) +
  xlab("number of leaves") +
  ylab("relative cost") +
  theme_bw()
```

We see a harsh decrease in the added cost after '1 leaf'.
Nevertheless, from visual inspection of the data we note that a backbone with three leaves may be more suitable.
We can increase the number of leaves up until the solution obtained from the last CLOF iteration, here 8-1 = 7.
Hence, to extract the solution for three leaves (which requires no additional computation), we can proceed as follows.

```{r, fig.width=4, fig.height=4}
BCB <- get_new_leaves(BCB, leaves=3)
EBCB <- get_edges2D(df, BCB$B)
ggplot(df[names(V(BCB$G)),], aes(x=x, y=y)) +
  geom_segment(data=EG, aes(x=x1, y=y1, xend=x2, yend=y2), color="black", alpha=0.1) +
  geom_segment(data=Epine, aes(x=x1, y=y1, xend=x2, yend=y2), color="black", alpha=0.5, size=1) +
  geom_point(size=3, fill=BCB$col, alpha=0.5, pch=21) +
  geom_segment(data=EBCB, aes(x= x1, y=y1, xend=x2, yend=y2), col="black", size=2.75) +
  geom_segment(data=cbind(EBCB, as.factor(BCB$branch)),
               aes(x=x1, y=y1, xend=x2, yend=y2, col=as.factor(BCB$branch)), size=2) +
  geom_point(data=df[V(BCB$B)$name,], fill=BCB$col[V(BCB$B)$name], size=5, pch=21) +
  scale_color_manual(values=BCB$palette) +
  guides(col = guide_legend(title="Path")) +
  coord_fixed() +
  theme_bw()
```

As mentioned, CLOF is standard solved for the betweenness centrality as a vertex-valued cost on the pine.
Nevertheless, we can use any nonnegative vertex-valued function for this purpose, such as the vertex degree.
Furthermore, since the proximity graph and boundary coefficients were already computed, we can immediately pass them as arguments to avoid unnecessary recomputation.
As such, we can conduct the backbone pipeline a second time for a different CLOF optimization function but the same graph/pine as follows.

```{r, fig.width=4, fig.height=4}
CLOF_vcost <- function(G) return(degree(G))
BCB2 <- backbone(BCB$G, f=BCB$f, leaves=3, CLOF_vcost=CLOF_vcost, assign=TRUE)
EBCB2 <- get_edges2D(df, BCB2$B)
ggplot(df[names(V(BCB$G)),], aes(x=x, y=y)) +
  geom_segment(data=EG, aes(x=x1, y=y1, xend=x2, yend=y2), color="black", alpha=0.1) +
  geom_segment(data=Epine, aes(x=x1, y=y1, xend=x2, yend=y2), color="black", alpha=0.5, size=1) +
  geom_point(size=3, fill=BCB2$col, alpha=0.5, pch=21) +
  geom_segment(data=EBCB2, aes(x= x1, y=y1, xend=x2, yend=y2), col="black", size=2.75) +
  geom_segment(data=cbind(EBCB2, as.factor(BCB2$branch)),
               aes(x=x1, y=y1, xend=x2, yend=y2, col=as.factor(BCB2$branch)), size=2) +
  geom_point(data=df[V(BCB2$B)$name,], fill=BCB2$col[V(BCB2$B)$name], size=5, pch=21) +
  scale_color_manual(values=BCB2$palette) +
  guides(col = guide_legend(title="Path")) +
  coord_fixed() +
  theme_bw()
```

Note that we now immediately specified that three leaves should be extracted, and no estimate is performed.
We can also use an edge-valued function to solve CLOF, such as the edge-weights.
However, note that in the weighted case, this function is non-constant on leaves.
Hence, no prepruning will be applied, significantly increasing the potential number of leaves and computational cost.
To (partially) accomodate for this, we can specify the exact (as above) or a maximal (as follows) allowed number of leaves.

```{r, fig.width=4, fig.height=4}
CLOF_ecost <- function(G) return(E(G)$weight)
BCB3 <- backbone(BCB$G, f=BCB$f, max_leaves=5, CLOF_ecost=CLOF_ecost, assign=TRUE)
EBCB3 <- get_edges2D(df, BCB3$B)
ggplot(df[names(V(BCB$G)),], aes(x=x, y=y)) +
  geom_segment(data=EG, aes(x=x1, y=y1, xend=x2, yend=y2), color="black", alpha=0.1) +
  geom_segment(data=Epine, aes(x=x1, y=y1, xend=x2, yend=y2), color="black", alpha=0.5, size=1) +
  geom_point(size=3, fill=BCB3$col, alpha=0.5, pch=21) +
  geom_segment(data=EBCB3, aes(x= x1, y=y1, xend=x2, yend=y2), col="black", size=2.75) +
  geom_segment(data=cbind(EBCB3, as.factor(BCB3$branch)),
               aes(x=x1, y=y1, xend=x2, yend=y2, col=as.factor(BCB3$branch)), size=2) +
  geom_point(data=df[V(BCB3$B)$name,], fill=BCB3$col[V(BCB3$B)$name], size=5, pch=21) +
  scale_color_manual(values=BCB3$palette) +
  guides(col = guide_legend(title="Path")) +
  coord_fixed() +
  theme_bw()
```

The specified max number of leaves (5) was below the total number of leaves (7) in the prepruned pine above.
Nevertheless, we still observe that it still takes relatively much longer to obtain the backbone.
This is because solving CLOF starts by computing the distance of all nodes in the pine to all leaves in the pine.
Both these numbers tend do decrease drastically when prepruning the pine, showing the usefulness of functions constant on leaves in conjunnction with CLOF.
Furthermore, as no exact number of leaves was specified, we see that this time the elbow estimate does lead to three leaves.
However, we observe that these are chosen rather randomly among the noise, again showing the usefulness of prepruning the pine.
Finally, now that we have three differently obtained backbones, we can evaluate and compare them as follows.

```{r}
backbones_evaluate(list(BCB, BCB2, BCB3))
```
